---
title: "Working with data"
format: 
  html:
    toc: true
    toc-depth: 3
editor: visual
---

# Brainstorm

-   Data frames. They're lists, but special

-   *Tabular data*

    -   What does good data look like? (get to this after showcasing some clean data)

-   Tidyverse - dplyr, readr, tidyr, etc

-   Basic verbs

    -   mutate - add/change column

    -   select - subset columns

    -   filter - subset rows

    -   group_by and summarize - aggregate

-   Utility verbs

    -   arrange - sorting

    -   distinct - unique combinations

    -   count - counts of combinations

-   Advanced verbs

    -   \[left,right,full\]\_join - joining tables

    -   \[semi,anti\]\_join - filter with tables

    -   pivot_longer

    -   pivot_wider

-   Pipes!

## Data

Small Mammal Exclosure Study (SMES) Vegetation Data from the Chihuahuan Desert Grassland and Shrubland at the Sevilleta National Wildlife Refuge, New Mexico (1995-2009)

Identifier: knb-lter-sev.97.182102

```{r}
#install.packages("dataone")
download.file("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sev.97.182102&entityid=79b6824dc326ff6b5b40eb4aa51a534a", 
              "data/sev097_smesquad_20201203.csv")
```

```{r}
library(here)
library(tidyverse)
smes_quads <- read_csv(here("data/sev097_smesquad_20201203.csv"))
```

USDA plant list

<https://plants.sc.egov.usda.gov/home/downloads>

```{r}
plantschecklist <- read_csv(here("data/completeplantschecklist.txt"))
```
